Outline flowCore paper

### 2008-07-16 paper/manuscript.tex
some updates to the paper manuscript
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@32728 bc3139a8-67e5-0310-9ffc-ced21a209358

\section*{Introduction}
During the last several years, automation technologies have been
developed that enable the use of flow cytometry high content screening
(FC-HCS) to generate large, complex datasets in both basic and
clinical research applications \citep{brinkman2007hcf}. A serious
bottleneck in the intepretation of existing studies and the
application of FC-HCS to even larger, more complex problems is that
data management and data analysis methods have not advanced
sufficiently far from the methods developed for traditional
applications of flow cytometry (FCM) to small-scale, tube-based
studies.

Some of the consequences of this lag in development are difficulties
in maintaining the integrity and documentation of extremely large
datasets, assessing measurement quality, developing validated assays,
assessing the accuracy of gating techniques, automating complex gating
strategies, and aggregating statistical results across large study
sets for further analysis.  Among the most serious consequences of the
current situation is, however, that it is very difficult for new
analysis approaches to find there way into standard practice. We
believe that this barrier to the development and dissemination of new
analysis methods is one of the fundamental restraints on the future
expansion of FC-HCS in both clinical and research applications.

@@

\subsection*{R and Bioconductor}
R is a robust statistical programing environment which offers a wide
range of statistical and visualization methods developed for various
fields of application. In particular, the Bioconductor project aims to
provide R software modules for biolocial and clinical data analysis.
It is widely used to process data in various applications, spanning a
diversity of research fields from genomics to proteomics to cell
biology \citep{gentleman2006bos}. The R language and Bioconductor are
particularly geared towards the analysis of large complex datasets
that typically arise from high throughput experiments such as
microarray gene expression analysis or FC-HCS. Using flow cytometry as
a high-throughput technology requires suitable software infrastructure
to facilitate the rapid handling of the data obtained. We rely on two
important lessons learned from the field of gene expression data
analysis: the first being the importance of data structures that
reflect the underlying data and facilitate the manipulations that are
of most interest, while the second lesson being the importance of a
modular architecture that allows for many developers to extend and use
the underlying infrastructure and to combine in complex workflows.

@@

\section*{Basic data structures}
\subsection*{flowFrame}
flowCore's primary task is the representation and basic manipulation
of FCM data. This is accomplished through a data model very close to
that adopted by other Bioconductor packages. All information from a
single FCS file, i.e., the collection of events and the accompanying
metadata, is stored in one single container. We call the structure
that hold this data a flowFrame, and raw data values as well as
associated metadata can be accessed programmatically. Most commonly,
these are descriptors of the stains used in the experiment and the
respective measurement channels, information about compensation
performed at the instrument side and any additional keywords the user
deems to be important to annotate the data. A number of quality checks
are performed during the creation of a flowFrame to ensure data
integrity.

\subsection*{flowSet}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flowFrames called a
flowSet. One can envision a flowSet to be a container for multiple
flowFrames along with relevant information associated with each
individual frame such as descriptions of the cell sample or the
treatment to which the sample was subjected. They manage the
consistent application of operations on the individual flow frames and
shift the burden of keeping score of the annotation from the user to
the infrastructure, thus reducing the risk of common errors
(\textit{e.g.}, mixups of sample labels). The flowSet structure can be
readily extended to incorporate the potentially complex annotation
associated with large FC-HCS experiments. Access to flowCore data
structures and methods has already enabled the development of new
Bioconductor packages, such as plateCore, that faciliate the analysis
of plate-based FC-HCS data and make handling the annotation and
processing the data easier.

@@

\section*{Standard flow operations}
The basic operations in FCM analyses are typically the same: the data
need to be compensated (if not already done on the instrument side)
and transformed, and sup-populations of interest need to be selected
based on a set of (predominantly sequential) gates. All software
solutions for FCM analysis offer more or less elaborate support for
these kinds of operation, and most of them in an interactive,
GUI-driven interface. The approach we have taken in flowCore is to
abstractly describe these operations and build a set of tools to
perform them on both flow frames and flow sets. While transformation,
and to a certain extent compensation, are fairly routine operations
with only limited potential for improvement, being able to implement
new methodologies for gating of FCM data, and extending the
capabilities of flowCore through object oriented programming are
features that clearly sets our framework apart from other FCM analysis
tools. By factoring out as much of the bookkeeping as possible,
programmers can focus on the actual operations rather then having to
deal with the tedious details of data integration and
access. Third-party methods can act on their own as first-class
citizens in the analysis framework, without breaking the workflow or
the basic infrastructure. This design allows for the straightforward
extension of flowCore's capabilities, and has already fostered the
development of a number of valuable add-ons
\citep{lo2008agf,sarkar2008ufv} \textit{(plateCore}.

\subsection*{Transformation and compensation}
Data transformation is essential for both data visualization and
modelling \citep{lo2008agf}. The major transformations that are
routinely used in FCM analysis have been implemented in flowCore
(\textit{e.g.,} quadratic, log or arcsinh). Furthermore, the design of the R language makes
it easy to define arbitrary functions to apply to the data of
individual flow frames or entire flow sets, respectively. Compensation
is available for both flow frames and flow sets, and the software also
offers functionality to compute spillover or compensation matrices
from a set of appropriate compensation samples.

\subsection*{Gating}
In flowCore, gating operations are represented by classes that can be
extended in an object-oriented manner. Basic gate types such as
rectangular gates, ellipses and polygon gates are implemented as part
of the framework. In addition, we introduce the notion of data-driven
gates, or filters, for which the necessary parameters are computed
based on the properties of the underlying data, for instance by
modelling underlying data distribution or by density estimation. The
ability to programmatically access gates is a prerequisite for
automated or semi-automated gating. By utilizing a unified interface
for all different types of gates, the user is able to subset data sets
as well as to create summary statistics, like the proportions of
events falling in single gates or in the combinations of
gates. Complex combinations and hierarchies of gates can be captured
in objects of class filterSet, allowing to apply multi-step gating
strategies. The definition of gates in flowCore follows the Gating-ML
CR, thus any flowCore gating strategy can be reproduced by any other
software that also adheres to the standard.

\section*{Related flow packages}
In addition to the flowCore package that offers basic infrastructure,
we have implemented a range of additional Bioconductor packages that
are dedicated to more specific tasks of FCM data analysis
(Figure \ref{fig2:flowWheel}). The flowViz package provides
sophisticated data visualization tools, that make use of multivariate
trellis plotting. These functions can be used to quickly generate
customized plots for extended cytometry data sets for both direct data
inspection and quality control. The flowQ package offers more advanced
quality assurance (QA) methodology and a framework to create
interactive web-based reports of QA results. Most of flowUtil's
content deals with data import and output from and to flowCore and the
flow-cytometry specific standard markup languages. Finally, flowStats
provides statistical methods that are relevant in the context of flow
cytometry data analysis.

@@

A very basic matrix of scatterplots is shown in Figure \ref{xyplot},
where each panel in the matrix represents the data of one individual
patient. Instead of plotting individual points, the local bivariate
density of the data is shown as a false color image. For each panel,
the outlines of the regions identified in the automated gating
operation described below are added to the plot. The input to the
plotting function is either a single flow frame or a complete flow set
object, and all available metadata can be accessed to control the
appearance and the composition of the plot. Similarly, gate objects
are directly passed on to the plotting function and are included in
the plot if possible. Besides the very basic scatterplot shown here,
there is a wealth of additional data visualizations available in the
flowViz package\cite{sarkar2008ufv}, all making use of flowCore's data
models allowing to easily arrange the data according to numerous
experimental factors. Furthermore, the design and the API of the
visualization software itself is very generic, and the user can
readily extend its capabilities by providing self-defined plotting
functions.

\begin{figure}[htbp]
\centering
\includegraphics[width=0.75\textwidth]{xyplot}
\caption{\label{xyplot}%
Scatterplot matrix of a single Flow Set with outlines of the gating
regions identified by an automated gating operation.}
\end{figure}

>>>>

\section*{Introduction}
\subsection*{Computational tools for the analysis of FC-HCS data}
During the last several years, automation technologies have been
developed that enable the use of flow cytometry high content screening
(FC-HCS) to generate large, complex datasets in both basic and
clinical research applications \citep{brinkman2007hcf}. A serious
bottleneck in the intepretation of existing studies and the
application of FC-HCS to even larger, more complex problems is that
data management and data analysis methods have not advanced
sufficiently far from the methods developed for traditional
applications of flow cytometry (FCM) to small-scale, tube-based
studies.

Some of the consequences of this lag in development are difficulties
in maintaining the integrity and documentation of extremely large
datasets, assessing measurement quality, developing validated assays,
controling the accuracy of gating techniques, automating complex gating
strategies, and aggregating statistical results across large study
sets for further analysis.  Among the most serious consequences of the
current situation is, however, that it is very difficult for new
analysis approaches to find there way into standard practice. We
believe that this barrier to the development and dissemination of new
analysis methods is one of the fundamental restraints on the future
expansion of FC-HCS in both clinical and research applications.

@@

\subsection*{R and Bioconductor}
R is a robust statistical programing environment which offers a wide
range of statistical and visualization methods developed for various
fields of application. In particular, the Bioconductor project aims to
provide R software modules for biolocial and clinical data analysis.
It is widely used to process such data, spanning a diversity of
research fields from genomics to proteomics to cell biology
\citep{gentleman2006bos}. The R language and Bioconductor are
particularly geared towards the analysis of large complex datasets
that typically arise from high throughput experiments such as
microarray gene expression analysis or FC-HCS. Using flow cytometry as
a high-throughput technology requires suitable software infrastructure
to facilitate the rapid handling of the data obtained. We rely on two
important lessons learned from the field of gene expression data
analysis: the first being the importance of data structures that
reflect the underlying data and facilitate the manipulations that are
of most interest, while the second lesson being the importance of a
modular architecture that allows for many developers to extend and use
the underlying infrastructure and to combine its tools in complex
workflows.

@@

\section*{Materials and Methods}
\subsection*{Basic data structures}
\subsubsection*{flowFrame}
FlowCore's primary task is the representation and basic manipulation
of FCM data. This is accomplished through a data model very close to
that adopted by other Bioconductor packages. All information from a
single FCS file, i.e., the collection of events and the accompanying
metadata, is stored in one single container. We call the structure
that hold this data a flowFrame. Raw data values as well as associated
metadata of a flowFrame can be accessed programmatically. Most
commonly, these are descriptors of the stains used in the experiment
and the respective measurement channels, information about
compensation performed at the instrument side and any additional
keywords the user deems to be important to annotate the data. A number
of quality checks are performed during the creation of a flowFrame to
ensure data integrity.

\subsubsection*{flowSet}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flowFrames called a
flowSet. One can envision a flowSet to be a container for multiple
flowFrames along with relevant information associated with each
individual frame such as descriptions of the cell sample or the
treatment to which the sample was subjected. FlowSets manage the
consistent application of operations on the individual flowFrames and
shift the burden of keeping score of the annotation from the user to
the infrastructure, thus reducing the risk of errors (\textit{e.g.},
mixups of sample labels). The flowSet structure can be readily
extended to incorporate the potentially complex annotation associated
with even larger FC-HCS experiments. Access to flowCore data
structures and methods has already enabled the development of new
Bioconductor packages, such as plateCore, that faciliate the analysis
of plate-based FC-HCS data and make handling the annotation and
processing the data easier.

@@

\subsection*{Standard flow operations}
The basic operations in FCM analyses are typically the same: the data
need to be compensated (if not already done on the instrument side)
and transformed, and sup-populations of interest need to be selected
based on a set of (predominantly sequential) gates. All software
solutions for FCM analysis offer more or less elaborate support for
these kinds of operation in an interactive, GUI-driven interface. The
approach we have taken in flowCore is to abstractly describe these
operations and build a set of tools to perform them on both flowFrames
and flowSets. While transformation, and to a certain extent
compensation, are fairly routine operations with only limited
potential for improvement, being able to implement new methodologies
for gating of FCM data, and extending the capabilities of flowCore
through object oriented programming are features that clearly sets our
framework apart from other FCM analysis tools. By factoring out as
much of the bookkeeping as possible, programmers can focus on the
actual operations rather then having to deal with the tedious details
of data integration and access. Third-party methods can act on their
own as first-class citizens in the analysis framework, without
breaking the workflow or the basic infrastructure. This design allows
for the straightforward extension of flowCore's capabilities, and has
already fostered the development of a number of valuable add-ons
\citep{lo2008agf,sarkar2008ufv} \textit{(plateCore)}.

\subsubsection*{Transformation and compensation}
Data transformation is essential for both data visualization and
modelling \citep{lo2008agf}. The major transformations that are
routinely used in FCM analysis have been implemented in flowCore
(\textit{e.g.,} quadratic, log or arcsinh, see Table~\ref{table1} for a
complete list). Furthermore, the design of the R language makes it
easy to define arbitrary functions to apply to the data of individual
flowFrames or entire flowSets, respectively. Compensation is
available for both flowFrames and flowSets, and the software also
offers functionality to compute spillover or compensation matrices
from a set of appropriate compensation samples.

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|}
\hline
\multicolumn{2}{|c|}{Data Transformations} \\
\hline
linear & $ax + b$ \\
quadratic & $ax^2 + bx + c$ \\
natural logarithm & $ln(x)(r/d)$ \\
logarithm & log(x)(r/d) \\
biexponential & need equation \\
logicle& need equation \\
truncate & $x_{x<=a} = a$ \\
scale & (x-a)/(b-a) \\
split-scale & need equation \\
arcsinh & $asinh(a + bx)+c$ \\
\hline
\end{tabular}
\caption{\label{table1}Data transformations implemented in flowCore.}
\end{center}
\end{table}

\subsubsection*{Gating}
In flowCore, gating operations are represented by classes that can be
extended in an object-oriented manner (Table~\ref{table2}). Basic gate
types such as rectangular gates, ellipses and polygon gates are
implemented as part of the framework. In addition, we introduce the
notion of data-driven gates, or filters, for which the necessary
parameters are computed based on the properties of the underlying
data, for instance by modelling data distribution or by density
estimation. The ability to programmatically access gates is a
prerequisite for automated or semi-automated gating. By utilizing a
unified interface for all different types of gates, the user is able
to subset data sets as well as to create summary statistics, for
instance the proportions of events falling in single gates or in the
combinations of gates. Complex combinations and hierarchies of gates
can be captured in objects of class filterSet, allowing to apply
multi-step gating strategies. The definition of gates in flowCore
follows the Gating-ML CR, thus any flowCore gating strategy can be
reproduced by any other software that also adheres to the standard.

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|}
\hline
\multicolumn{2}{|c|}{Gates} \\
\hline
rectangleGate & n-dinensional rectangular regions \\
quadGate & quadrant regions in two dimensions \\
polygonGate & polygonal regions in two dimensions \\
polytopeGate & generalization of polygon in n dimensions \\
ellipsoidGate & n-dimesnional ellipsoid region \\
\hline
\multicolumn{2}{|c|}{Filters} \\
\hline
sampleFilter & random subsampling\\
expressionFilter & filter by resuts of a boolean expression \\
kmeansFilter & based on K-means clustering \\
norm2Filter & based on bivariate normal distribution \\
curv1Filter & based on local density regions in 1D \\
curv2Filter & based on local density regions in 2D \\
timeFilter & based on abnormal data acquisition over time \\
\hline
filterSet & gating strategy \\
\hline
\end{tabular}
\caption{\label{table2}Filter and gate classes implemented in flowCore.}
\end{center}
\end{table}

\subsection*{Related flow packages}
In addition to the flowCore package that offers basic infrastructure,
we have implemented a range of additional Bioconductor packages that
are dedicated to more specific tasks of FCM data analysis
%(Figure \ref{fig2:flowWheel}) 
. The flowViz package provides
sophisticated data visualization tools, that make use of multivariate
trellis plotting. These functions can be used to quickly generate
customized plots for extended cytometry data sets for both direct data
inspection and quality control. The flowQ package offers more advanced
quality assurance (QA) methodology and a framework to create
interactive web-based reports of QA results. Most of flowUtil's
content deals with data import and output from and to flowCore and the
flow-cytometry specific standard markup languages. Finally, flowStats
provides statistical methods that are relevant in the context of flow
cytometry data analysis.

@@

A very basic matrix of scatterplots is shown in Figure \ref{xyplot},
where each panel in the matrix represents the data of one individual
patient. Instead of plotting individual points, the local bivariate
density of the data is shown as a false color image. For each panel,
the outlines of the regions identified in the automated gating
operation described below are added to the plot. The input to the
plotting function is either a single flowFrame or a complete flowSet
object, and all available metadata can be accessed to control the
appearance and the composition of the plot. Similarly, gate objects
are directly passed on to the plotting function and are included in
the plot if possible. Besides the very basic scatterplot shown here,
there is a wealth of additional data visualizations available in the
flowViz package\cite{sarkar2008ufv}, all making use of flowCore's data
models allowing to easily arrange the data according to numerous
experimental factors. Furthermore, the design and the API of the
visualization software itself is very generic, and the user can
readily extend its capabilities by providing self-defined plotting
functions.

\begin{figure}[htbp]
\centering
\includegraphics[width=0.75\textwidth]{xyplot}
\caption{\label{xyplot}%
Scatterplot matrix of a single FlowSet with outlines of the gating
regions identified by an automated gating operation.}
\end{figure}
